A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-29 paper.tex
Lots of editing

\documentclass{llncs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fixltx2e}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{positioning}

@@

In this paper, we extend the method by Narodytska et al.~\cite{narodytska2014} to unbounded 
safety games.  To this end we enhance the method with computational learning: every time
the search algorithm discovers a new counterexample, the learning procedure analyzes the 
counterexample, extracting a subset of states winning for one of the players from it.  
The learning procedure ensures that reaching a fixed point in these sets is sufficient to establish 
unbounded realizability.
Our method can be seen as a hybrid between the counterexample-guided algorithm by
Narodytska et al. and methods based on winning set computation: we use the former to guide 
the search, while we rely on the latter to ensure convergence.

@@

Section 2 outlines the original bounded synthesis algorithm. In Section 3 we
describe and prove the correctness of our extension of the algorithm to
unbounded games. In the following sections we discuss optimisations to the
algorithm, evaluate our methodology, and compare our approach to other
synthesis techniques.

@@

At every round of the game, the \emph{environment} picks an uncontrollable
action, the \emph{controller} responds by choosing a controllable action and
the game transitions into a new state according to $\delta$.  A \emph{run} of a
game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n, c_n)$ is a chain of
state and action pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$.  A run is
winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot
E(x_i))$.  In a \emph{bounded game} with bound $\kappa$ all runs are restricted
to length $\kappa$, whereas unbounded games consider runs of infinite length.
Since we consider only deterministic games, a run is uniquely described by a
list of assignments to $U$ and $C$.

A \emph{controller strategy} $\pi^c : (X, U) \to C$ is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of rank $n$ if all runs $(x_0, u_0, \pi^c(x_0, u_0)), (x_1,
u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are winning.
Bounded \emph{realizability} is the problem of determining the existence of
such a strategy for a bounded game.

@@

Realisability for a bounded safety game can be solved by checking a quantified
formula consisting of the game's transition relation \emph{unrolled} to the
length of the bounded game. The quantifiers represent the two players of the
game: universal for the controller and existential for the environment.  Just
as the players alternate in choosing actions every step of the game, the
formula is prefixed by a quantifier alternation over $U$ and $C$ for each
step.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

@@

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, s, k, T, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$s, k, cand, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $k'$, $s'$, $\emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The algorithm takes a concrete game $G$ with maximum bound $\kappa$ as an implicit
argument.  In addition, it takes a player $p$ (controller or environment),
state $s$, bound $k$ and an abstract game tree $T$ and returns a winning
partial strategy for $\textsc{opponent}(p)$, if one exists.  The initial
invocation of the algorithm takes the initial state $I$, bound $\kappa$ and an
empty abstract game tree $\emptyset$.  Initially the solver is playing on
behalf of the environment since that player takes the first move in every step.
The empty game tree does not constrain opponent moves, hence solving such an
abstraction is equivalent to solving the original concrete game.

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}. Then copies of state and action variables are introduced for
each node in tree and opponent action variables for each edge.  Opponent
actions are set to $\textsc{action}(e)$, the action that labels the edge of the
tree $e$. The new state and action variables are connected to the successor
nodes (\textsc{succ}) by an encoding transition relation and constrained to the
winning condition of the player.

@@

We instead present an approach that iteratively solves games of increasing
bound while learning states. We show that reaching a fixed point in learned
states is sufficient for completeness.

@@

\begin{algorithm}[h]
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkRank}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
    \end{algorithmic}
    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

%%%The unbounded synthesis algorithm enhances the bounded method with
%%%computational learning. States are carefully recorded from unsatisfiable
%%%abstract games while maintaining an invariant ensuring the monotonicity of
%%%learning.

@@

We modify the bounded synthesis algorithm to learn a subset of states from
failed attempts to find candidate strategies.

When the abstract game is unsatisfiable this indicates that the states
represented by nodes in the abstract game tree are losing for the current
player. We extract these states from the game tree using interpolation.

Given two formulas $F_1$ and $F_2$ such that $F_1 \land F_2$ is unsatisfiable,
it is possible to construct a Craig interpolant\cite{craig1957} $\mathcal{I}$
such that $F_1 \to \mathcal{I}$, $F_2 \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $F_1$ and $F_2$.
An interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $F_1 \land F_2$ \cite{pudlak1997}.

\begin{figure}
    \centering
    \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
        \node [circle,draw] (root){}
            child {node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$b$}
                }
                edge from parent [fixed] node [left] {$a$}
            }
            child {node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$e$}
                }
                edge from parent [fixed] node [right] {$d$}
            }
            node [above=4pt] {$\langle s, 2 \rangle$};
    \end{tikzpicture}
    \caption{A controller-losing game tree}
    \label{fig:interpolatetree}
\end{figure}

Consider the game tree in Figure \ref{fig:interpolatetree}. The
tree is losing for the controller, the topmost node is at rank 1, and the $a$ and
$b$ nodes are at rank 0. Since the tree is controller-losing we know that at
least one state in the tree will uncontrollably lead to the error state.  As a
result, $\textsc{treeFormula}(T)$ is unsatisfiable. If we take a step of the
tree, such as the step where the environment plays $a$, and cut it from the
rest of the tree then $\textsc{treeFormula}(step) \land
\textsc{treeFormula}(parent)$ must too be unsatisfiable.

We can construct an interpolant with $F_1 = \textsc{treeFormula}(parent)$ and
$F_2 = \textsc{treeFormula}(step)$. The only variables shared between $F_1$ and
$F_2$ are the state variables at rank 1. We know that $F_2 \land \mathcal{I}$
is unsatisfiable, therefore all states in $\mathcal{I}$ must lose to the
uncontrollable action $a$. We also know that $F_1 \to \mathcal{I}$, thus
$\mathcal{I}$ contains all states reachable by the parent tree (on runs that
avoid the error state.)

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment respectively}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{tf2}
        \Function{\textoverline{treeFormula}}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{\Call{E}{$x^{gt}$}}
        \Else
        \State \Return{ $\Call{$B^m[rank(gt)]$}{x^{gt}} \land$ \\
            $(\Call{E}{x^{gt}} \lor \bigvee_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{$\overline{treeFormula}$}{n}))$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Now we can consider the step where the environment plays $b$, and the parent
now without either step in the snippet. The formula
$(\textsc{treeFormula}(parent) \land \mathcal{I}) \land
\textsc{treeFormula}(step)$ must be unsatisfiable.  $\mathcal{I}$ contains all
states that lose to $a$ so any other state reachable at rank 1 in this subtree
must lose to $b$.  Therefore we can compute another interpolant that contains
states that lose to $b$.

This is the foundation for a recursive algorithm that consumes an entire tree
by removing a single step on each iteration (Algorithm \ref{alg:learn}). All
learned states from which the controller must lose are recorded in a set of bad
states $B^M$. Environment-losing are learned in the same way but are recorded
by removing the state from a set of sets of controller may-losing states $B^m$.
An environment-losing state of rank $n$ is removed from every set at rank $n$
and lower.

We compile interpolants into DNF cubes by constructing a temporary BDD. This
smaller representation is necessary for maintaining the size of formulas passed
to the SAT solver. However, it can occaisionally lead to state explosion.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Require $\sigma(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $\sigma(X_T) \land B^M$
        \Function{learn}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with max rank
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$\sigma, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $\sigma(X_T) \land B^m[$\Call{rank}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with max rank
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{rank}{$n$}}
                \State $B^m[i] \gets B^m[i] \land \lnot \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$\sigma, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

The correctness of the unbounded synthesis algorithm can be established
independently from that of the bounded algorithm. Fortunately, the correctness
of the bounded solver has been established in \cite{narodytska2014}.

We define two global invariants of the algorithm.  The \emph{may-invariant}
states that sets $B^m[i]$ grows monotonically with $i$ and that each $B^m[i+1]$
overapproximates the uncontrollable predecessor of $B^m[i]$: $$\forall
i<k.~B^m[i] \subseteq B^m[i+1], Upre(B^m[i]) \subseteq B^m[i+1].$$

@@

The sets $B^m$ and $B^M$ are only modified by the inductive solver, implemented
by \textsc{learn} and \textoverline{\textsc{learn}} functions.  Below we prove that these
functions indeed maintain the invariants.

\subsection{Proof of \textsc{\textoverline{learn}}}

@@

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is
of depth 1, this means that the environment cannot force the game into
$B^m[\textsc{rank}(n)]$ playing against the counterexample moves in $T_2$.
Hence, $\II \cap Upre(B^m[\textsc{rank}(n)]) = \emptyset$.  Furthermore, since
the may-invariant holds, $\II \cap Upre(i) = \emptyset$, for all $i <
\textsc{rank}(n)$.  Hence, removing $\II$ from all $B^m[i], i\leq
\textsc{rank}(n)$ in line~(15) preserves the may-invariant, thus satisfying the
first post-condition.

>>>>

\documentclass{llncs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fixltx2e}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{positioning}
\usetikzlibrary{plotmarks}

@@

In this paper, we extend the method by Narodytska et al.~\cite{narodytska2014} to unbounded 
safety games.  To this end we enhance the method with computational learning: every time
the search algorithm discovers a new counterexample, the learning procedure analyzes the 
counterexample, extracting a subset of states winning for one of the players from it.  
The learning procedure ensures that reaching a fixed point in these sets is sufficient to establish 
unbounded realizability.
Our method can be seen as a hybrid between the counterexample-guided algorithm
by Narodytska et al. and methods based on losing set computation: we use the
former to guide the search, while we rely on the latter to ensure convergence.

@@

Section 2 outlines the original bounded synthesis algorithm. In Section 3 we
describe and prove the correctness of our extension of the algorithm to
unbounded games. In the following sections we evaluate our methodology, and
compare our approach to other synthesis techniques.

@@

At every round of the game, the \emph{environment} picks an uncontrollable
action, the \emph{controller} responds by choosing a controllable action and
the game transitions into a new state according to $\delta$.  A \emph{run} of a
game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n, c_n)$ is a chain of
state and action pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$.  A run is
winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot
E(x_i))$.  In a \emph{bounded game} with maximum bound $\kappa$ all runs are
restricted to length $\kappa$, whereas unbounded games consider runs of
infinite length.  Since we consider only deterministic games, a run is uniquely
described by a list of assignments to $U$ and $C$.

A \emph{controller strategy} $\pi^c : (X, U) \to C$ is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of maximum bound $\kappa$ if all runs $(x_0, u_0, \pi^c(x_0,
u_0)), (x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are
winning.  Bounded \emph{realizability} is the problem of determining the
existence of such a strategy for a bounded game.

@@

Realisability for a bounded safety game can be solved by checking a quantified
formula consisting of the game's transition relation \emph{unrolled} to the
length of the bounded game. The quantifiers represent the two players of the
game: universal for the controller and existential for the environment.  Just
as the players alternate in choosing actions every round of the game, the
formula is prefixed by a quantifier alternation over $U$ and $C$ for each
round.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine $T'$ with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined game tree}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

@@

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, s, k, T, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$s, k, cand, l$} \Comment{Get bound and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $k'$, $s'$, $\emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The algorithm takes a concrete game $G$ with maximum bound $\kappa$ as an implicit
argument.  In addition, it takes a player $p$ (controller or environment),
state $s$, bound $k$ and an abstract game tree $T$ and returns a winning
partial strategy for $\textsc{opponent}(p)$, if one exists.  The initial
invocation of the algorithm takes the initial state $I$, bound $\kappa$ and an
empty abstract game tree $\emptyset$.  Initially the solver is playing on
behalf of the environment since that player takes the first move in every game
round.  The empty game tree does not constrain opponent moves, hence solving
such an abstraction is equivalent to solving the original concrete game.

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}. Then copies of state and action variables are
introduced for each node in tree and opponent action variables for each edge.
Opponent actions are set to $\textsc{action}(e)$, the action that labels the
edge of the tree $e$. The state and action variables of node $n$ are connected
to successor nodes $\textsc{succ}(n)$ by an encoding of the transition relation
and constrained to the winning condition of the player.

@@

We instead present an approach that iteratively solves games of increasing
bound while learning states with interpolation over abstract games. We show
that reaching a fixed point in learned states is sufficient for completeness.

@@

\begin{algorithm}[h]
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets \top$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkBound}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkBound}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
    \end{algorithmic}
    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

@@

We modify the bounded synthesis algorithm to learn a subset of states from
failed attempts to find candidate strategies. When \textsc{findCandidate}
cannot find a strategy for an abstract game tree beginning at a state $s$, we
know that $s$ is a losing state. We can also attempt to learn additional losing
states from the tree via interpolation.

\begin{figure}
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$b$}
                };
%%%                node [above=4pt] {$\langle s', 1 \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
\end{figure}

Consider the controller-losing game tree, $T$, in
Figure~\ref{fig:interpolatetree}. We choose any leaf node $l$ of $T$ and its
parent $n$. We then split the tree at $n$ such that both slices $T_1$ and $T_2$
contain $n$. Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$
except $l$, and $T_2$ (Figure~\ref{fig:treef2}) contains only $n$ and $l$ - a
single round of the game.  There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(T)$ is unsatisfiable, and by extension $s \land
\textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$ must also be
unsatisfiable. This enables the construction of an \emph{interpolant} that
captures losing states at node $n$.

Given two formulas $F_1$ and $F_2$ such that $F_1 \land F_2$ is unsatisfiable,
it is possible to construct a Craig interpolant~\cite{craig1957} $\mathcal{I}$
such that $F_1 \to \mathcal{I}$, $F_2 \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $F_1$ and $F_2$.
An interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $F_1 \land F_2$ \cite{pudlak1997}.

We can construct an interpolant with $F_1 = s \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$. The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By construction
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
must lose to the action in the edge between $n$ and $l$, the edge labelled by
action $b$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding the error state.

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{$B^M$}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{\Call{E}{$X_T$}}
        \Else
        \State \Return{ \Call{$B^m[\Call{bound}{T}]$}{$X_{T}$} $\land$ \\
            $$\Call{E}{X_T} \lor \bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

When we discover losing states for the controller we record them in a set of
states that \emph{must} be bad, $B^M$. This set is initialised to the error
state at the beginning of the algorithm and grows as losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (see Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with
$B^M$. This enables further interpolants to be constructed by recursively
splitting more nodes from $T_1$ since the states that lost to $T_2$ are now
blocked by $B^M$.

It is more complicated to learn states from environment-losing abstract game
trees. States are only losing for the environment for a particular bound. For a
state learned at a node $n$ it is losing at $\textsc{bound}(n)$. Therefore we
maintain an array of state sets indexed by the bound at which that set of
states is losing for the environment. Instead of accumulating states, as we do
in $B^M$, we instead remove states from the array $B^m$. This array is
initialised with $\top$ and contains only states that \emph{may} lose for the
controller. The modified \textsc{\textoverline{treeFormula}} function
constrains the environment to may-losing states, which also allows for
recursive learning over the entire tree.

We compile interpolants into DNF cubes by constructing a temporary BDD. This
smaller representation is necessary for maintaining the size of formulas passed
to the SAT solver. However, it can occasionally lead to state explosion.

The recursive learning algorithms, \textsc{learn} (for the controller) and
\textsc{\textoverline{learn}} (for the environment), are detailed in
Algorithm~\ref{alg:learn}. When executing \textsc{findCandidate} in the
unbounded solver, lines 19 and 20 become active and call into the learning
procedures. The unbounded algorithm performs bounded synthesis for iteratively
increasing maximum bounds until the initial state is in $B^M$ (environment
wins) or $B^m$ reaches a fixed point (controller wins). We prove correctness in
the next section.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Require $\sigma(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $\sigma(X_T) \land B^M$
        \Function{learn}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$\sigma, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $\sigma(X_T) \land B^m[$\Call{bound}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{bound}{$n$}}
                \State $B^m[i] \gets B^m[i] \land \lnot \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$\sigma, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

We define two global invariants of the algorithm.  The \emph{may-invariant}
states that sets $B^m[i]$ grows monotonically with $i$ and that each $B^m[i+1]$
overapproximates the states from which the environment can force the game into
$B^m[i]$. We call this operation $Upre$, the uncontrollable predecessor. So the
\emph{may-invariant} is: $$\forall i<k.~B^m[i] \subseteq B^m[i+1], Upre(B^m[i])
\subseteq B^m[i+1].$$

@@

The sets $B^m$ and $B^M$ are only modified by the functions \textsc{learn} and
\textoverline{\textsc{learn}}.  Below we prove that these functions indeed
maintain the invariants.

\subsection{Proof of learning functions}

@@

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is
of depth 1, this means that the environment cannot force the game into
$B^m[\textsc{bound}(n)]$ playing against the counterexample moves in $T_2$.
Hence, $\II \cap Upre(B^m[\textsc{bound}(n)]) = \emptyset$.  Furthermore, since
the may-invariant holds, $\II \cap Upre(i) = \emptyset$, for all $i <
\textsc{bound}(n)$.  Hence, removing $\II$ from all $B^m[i], i\leq
\textsc{bound}(n)$ in line~(15) preserves the may-invariant, thus satisfying the
first post-condition.

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{bound}(n)]$ makes $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.
